The Success of College Basketball Teams can be predicted using several advanced team statistics

### 2022-04-26 Paper-On-NCAAB-Stats/outputs/paper/Paper-On-NCAAB-Stats.Rmd
Completed writing

#Load in libraries
library(tidyverse)
library(dplyr)
library(knitr)
library(kableExtra)
library(ggpubr)
#install.packages("modelsummary")
library(modelsummary)
#install.packages("DiagrammeR")
library(DiagrammeR)
#install.packages("tidymodels")
library(tidymodels)
#install.packages("broom")
library(broom)
library(plot3D)
```

@@

March is a very exciting time for sports fans in the US and many other countries, as it is when the NCAA Men's Division 1 Basketball Tournament, more commonly known as March Madness, is played. 68 college basketball teams go head to head, playing single elimination games until there is only one team remaining, making for some of the most intense and unpredictable sports games in the world. As a result of this, March Madness draws in a lot of viewers. 16.9 million people tuned in to the 2021 March Madness finals, making it the 2nd most viewed North American sports event only behind the Superbowl [@citeBaker]. A big reason for the massive following is that many people create brackets for the tournament, trying to predict the results. With 9.2 quintillion different bracket possibilities, all the insane upsets that nobody sees coming, and the overall chaotic nature of the tournament, creating a bracket and seeing how well you do is very appealing to both sports fans and regular people, causing all the hype around March Madness [@citeGeiling]. People put a lot of work into their brackets, comparing teams to one another and trying to figure out who has the best odds of winning. With this in mind, is there a way fans can more successfully predict which teams will have more success than others in the tournament?

In this paper, we attempt to answer this intriguing question by analyzing a variety datasets from the college basketball section of Sports Reference [@citeSR], using R [@citeR]. Datasets containing college basketball statistics from 2017 to 2020 were used, where each dataset contains many different advanced team statistics for all NCAA Division 1 schools with a Men's basketball team in a specific season. The data was first graphically analyzed to determine which team statistics have a strong or noticeable relationship with how successful a team is in their games. These statistics were then used to create a linear model that attempts to predict a teams Simple Rating System, the response variable that is being used to measure team success, where only the best and most necessary team statistics were included. As a result of this, it was determined that the following statistics [Insert statistics here once model is completed], play the most important factor in leading to team success. Thus, when filling out their brackets, fans can compare these statistics across teams to determine which teams have a good chance of going far or even winning the tournament, as opposed to just looking at wins or their seeding like many people do.

@@

```{r scatters, echo=FALSE, fig.width=12, fig.height=12, fig.cap="Relationship between Simple Rating System and several adcanced team statistics, coloured by whether a schools winning % was above .500 or not", fig.pos="H"}

@@

```{r histograms, echo=FALSE, fig.cap="Distributions for important team statistics that affect a college basketball teams success", fig.pos="H"}

@@

```
Figure \@ref(fig:histograms) shows the distributions for each of these important team statistics that may be used as predictor variables, which allows us to more easily and visually see what was learned in the summary table. It can be seen that each variable seems to very closely resemble normal distributions, where they are centered around their mean and are fairly symmetric on both sides of it. Of course they are not perfectly symmetric as that would be near impossible, but the overall shape of the distributions are still very similar to that of a normal distribution. Offensive rating and total rebound percentage have a slight tail to the left of the distributions skewing them slightly to the left, and true shooting percentage and turnover percentage are the opposite where they are slightly skewed to the right due to their tails. However, they tails are quite small and not too noticeable, so the distributions on not skewed very much, which is evident by the means and medians of each variable being the exact same or nearly the same. Additionally, although there are some points at the end of the tails that may alter the distribution slightly, we can now more clearly see that there does not seem to be any substantially large outliers for any of the variables that would skew results too drastically. All in all, these team statistics should make for great explanatory variables and should not cause any issues in our model.

@@

DiagrammeR::grViz("
digraph {
  graph [ranksep = 0.2]
  node [shape = plaintext, fontsize = 10, fontname = Helvetica]
    A [label = 'True Shooting % / Total Rebound % / Turnover %']
    B [label = 'Offensive Rating']
    C [label = 'Simple Rating System']
  edge [minlen = 2, arrowhead = vee]
    A -> C
    A -> B
    B->C
  { rank = same; A; C}
}
", height = 300)
```
In the DAG, it can be seen how true shooting percentage, total rebound percentage, and turnover percentage all affect and have causality with simple rating system. It can also be seen that offensive rating affects simple rating system too, but it is also affected by the three variables listed above and is being used as a mediator between these variables and simple rating system. As this mediator, offensive rating is explaining exactly why these three variables affect simple rating system [@citeBhandari]. The three variables affect team success because the three variables cause teams to either score more or less points on offense, giving them either a better or worse offensive rating, and thus causing them to either win or lose more games, which is essentially what the DAG is explaining. As a result of this, offensive rating should not be used in the model as the mediator will affect the strength of the relationship between the true predictor variables and simple rating system.

@@

```{r coefsum}
final_model <- lm(srs~ts_percent + trb_percent + tov_percent, data=ncaab_stats_train)

@@

kable(sum_table,
    caption= "Summary of parameter values and statistics for linear model predicting Simple Rating System using variables True Shooting %, Total Rebound %, and Turnover %",
    booktabs = TRUE,
    linesep = "",
    align = c('c', 'c', 'c')
  ) %>% 
kable_styling(latex_options= "hold_position")

@@

We can plug all these coefficients into the $\beta's$ in the formula explained in Section \@ref(model) to get the completed formula for our final model, $\hat{y_i} = -98.62 + 0.92x_1 + 1.63x_2 - 2.04x_3$. Doing so makes our model much easier to interpret. In order to estimate what a teams simple rating system value would be, you plug in their true shooting percentage for $x_1$, their total rebound percentage for $x_2$, and their turnover percentage for $x_3$, and the output of $\hat{y_i}$ would be their rating. For example, if a team has a true shooting percentage of 50%, a total rebounding percentage of 40%, and a turnover percentage of 20%, the model estimates that their simple rating system would be $\hat{y_i} = -98.62 + 0.92(50) + 1.63(40) - 2.04(20) = -28.22 $. This shows how the model can successfully predict a teams success with the use of advanced team statistics such as our predictor variables that were used, answering the research question.

@@

scatter3D(ncaab_stats_train$ts_percent, ncaab_stats_train$trb_percent, ncaab_stats_train$srs, colvar = ncaab_stats_train$tov_percent, bty = 'b2',
          pch = 19, cex = 0.5, xlab = "TS%", ylab = "TRB%", zlab = "SRS", clab = "TOV%", ticktype = "detailed", )
```
Figure \@ref(fig:3d) shows this relationship through the use of a 3D plot, which is technically a 4D plot based on the use of a colour gradient to show a 4th variable. The vertical z-axis represents our response variable SRS, the x-axis closest to the viewer represents true shooting percentage, and the remaining y-axis on the right represents total rebound percentage. Also, the colour gradient is used to represent turnover percentage where each point is coloured in differently based on its turnover percentage value, giving the plot its 4th dimension.

@@

Finally, our model must be validated using the testing dataset. One way to do this is by making predictions on the test dataset, using the values of the explanatory variables in the data. We then calculate the RMSE using the errors between the predicted values and actual values of the testing data, and compare it to the RMSE of the original model built with the training data []. After using the model to make predictions on the testing data, it was found that the predicted values had a RMSE of 7.74. This is very similar to the original models RMSE based on the training data, which was 7.29. Since these values are very similar to one another, it means that our model did a good job at predicting values using the testing dataset, and thus it should also successfully be able to make predictions on other datasets, like for example future college basketball seasons. This means that the model has successfully been validated using the testing dataset, and we can now analyze what exactly was learned from our model.

@@

## First discussion point

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.

## Second discussion point

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.

\appendix

# Appendix {-}

# Additional details

>>>>

#Load in libraries
library(tidyverse)
library(dplyr)
library(knitr)
library(kableExtra)
#install.packages("ggpubr")
library(ggpubr)
#install.packages("modelsummary")
library(modelsummary)
#install.packages("DiagrammeR")
library(DiagrammeR)
#install.packages("tidymodels")
library(tidymodels)
#install.packages("broom")
library(broom)
#install.packages("plot3D")
library(plot3D)
```

@@

March is a very exciting time for sports fans in the US and many other countries, as it is when the NCAA Men's Division 1 Basketball Tournament, more commonly known as March Madness, is played. 68 college basketball teams go head to head, playing single elimination games until there is only one team remaining, making for some of the most intense and unpredictable sports games in the world. As a result of this, March Madness draws in a lot of viewers. 16.9 million people tuned in to the 2021 March Madness finals, making it the 2nd most viewed North American sports event only behind the Superbowl [@citeBaker]. A big reason for the massive following is that many people create brackets for the tournament, trying to predict the results. With 9.2 quintillion different bracket possibilities, all the insane upsets that nobody sees coming, and the overall chaotic nature of the tournament, creating a bracket and seeing how well you do is very appealing to both sports fans and regular people, causing all the hype around March Madness [@citeGeiling]. People put a lot of work into their brackets, comparing teams to one another and trying to figure out who has the best odds of winning. With this in mind, is there a way people can more successfully predict which teams will have more success than others in the tournament?

In this paper, we attempt to answer this intriguing question by analyzing a variety datasets from the college basketball section of Sports Reference [@citeSR], using R [@citeR]. Datasets containing college basketball statistics from 2017 to 2022 were used, where each dataset contains many different advanced team statistics for all NCAA Division 1 schools with a Men's basketball team in a specific season. The data was first graphically analyzed to determine which team statistics have a strong or noticeable relationship with how successful a team is in their games. These statistics were then used to create a linear model that attempts to predict a teams Simple Rating System, the response variable that is being used to measure team success, where only the best and most necessary team statistics were included. As a result of this, it was determined that the following statistics, true shooting percentage, total rebound percentage, and turnover percentage, play the most important factor in leading to team success. Using estimated coefficients, values of these statistics could be used to predict what a teams simple rating system value will be and how successful the team should be. Thus, when filling out their brackets, fans can compare these statistics across teams to determine which teams have a good chance of going far or even winning the tournament, as opposed to just looking at wins or their seeding like many people do.

@@

```{r scatters, echo=FALSE, fig.width=10, fig.height=10, fig.cap="Relationship between Simple Rating System and several adcanced team statistics, coloured by whether a schools winning % was above .500 or not", fig.pos="H"}

@@

```{r hists, echo=FALSE, fig.cap= "Distributions for important team statistics that affect a college basketball teams success"}

@@

```
Figure \@ref(fig:hists) shows the distributions for each of these important team statistics that may be used as predictor variables, which allows us to more easily and visually see what was learned in the summary table. It can be seen that each variable seems to very closely resemble normal distributions, where they are centered around their mean and are fairly symmetric on both sides of it. Of course they are not perfectly symmetric as that would be near impossible, but the overall shape of the distributions are still very similar to that of a normal distribution. Offensive rating and total rebound percentage have a slight tail to the left of the distributions skewing them slightly to the left, and true shooting percentage and turnover percentage are the opposite where they are slightly skewed to the right due to their tails. However, they tails are quite small and not too noticeable, so the distributions on not skewed very much, which is evident by the means and medians of each variable being the exact same or nearly the same. Additionally, although there are some points at the end of the tails that may alter the distribution slightly, we can now more clearly see that there does not seem to be any substantially large outliers for any of the variables that would skew results too drastically. All in all, these team statistics should make for great explanatory variables and should not cause any issues in our model.

@@

grViz(
"digraph {
  graph [ranksep = 0.2]
  node [shape = plaintext, fontsize = 10, fontname = Helvetica]
    A [label = 'True Shooting % / Total Rebound % / Turnover %']
    B [label = 'Offensive Rating']
    C [label = 'Simple Rating System']
  edge [minlen = 2, arrowhead = vee]
    A -> C
    A -> B
    B->C
  { rank = same; A; C}
}
", height = 300)
```
In the DAG, it can be seen how true shooting percentage, total rebound percentage, and turnover percentage all affect and have causality with simple rating system. It can also be seen that offensive rating affects simple rating system too, but it is also affected by the three variables listed above and is being used as a mediator between these variables and simple rating system. As this mediator, offensive rating is explaining exactly why these three variables affect simple rating system [@citeBhandari]. The three variables affect team success because the three variables cause teams to either score more or less points on offense, giving them either a better or worse offensive rating, and thus causing them to either win or lose more games, which is essentially what the DAG is explaining. As a result of this, offensive rating should not be used in the model as the mediator will affect the strength of the relationship between the true predictor variables and simple rating system.

@@

```{r coefsum, echo=FALSE}
final_model <- lm(srs~ts_percent + trb_percent + tov_percent, data=ncaab_stats_train)

@@

kable(sum_table,
    caption = "Summary of parameter values and statistics for linear model predicting Simple Rating System using various team statistics",
    booktabs = TRUE,
    linesep = "",
    align = c('c', 'c', 'c')
  ) %>% 
kable_styling(latex_options= "hold_position")

@@

We can plug all these coefficients into the $\beta's$ in the formula explained in Section \@ref(model) to get the completed formula for our final model, $\hat{y_i} = -98.62 + 0.92x_1 + 1.63x_2 - 2.04x_3$. Doing so makes our model much easier to interpret. In order to estimate what a teams simple rating system value would be, you plug in their true shooting percentage for $x_1$, their total rebound percentage for $x_2$, and their turnover percentage for $x_3$, and the output of $\hat{y_i}$ would be their rating. For example, if a team has a true shooting percentage of 50%, a total rebounding percentage of 40%, and a turnover percentage of 20%, the model estimates that their simple rating system would be $\hat{y_i} = -98.62 + 0.92(50) + 1.63(40) - 2.04(20) = -28.22$. This shows how the model can successfully predict a teams success with the use of advanced team statistics such as our predictor variables that were used, answering the research question.

@@

scatter3D(ncaab_stats_train$ts_percent, ncaab_stats_train$trb_percent, ncaab_stats_train$srs, 
          colvar = ncaab_stats_train$tov_percent, 
          bty = 'b2', pch = 19, cex = 0.5, 
          xlab = "TS%", ylab = "TRB%", zlab = "SRS", 
          clab = "TOV%", ticktype = "detailed")
```
Figure \@ref(fig:3d) shows this relationship through the use of a 3D plot, which is technically a 4D plot based on the use of a colour gradient to show a 4th variable. The vertical z-axis represents our response variable SRS, the x-axis closest to the viewer represents true shooting percentage, and the remaining y-axis on the right represents total rebound percentage. Also, the colour gradient is used to represent turnover percentage where each point is coloured in differently based on its turnover percentage value, giving the plot its 4th dimension.

@@

Finally, our model must be validated using the testing dataset. One way to do this is by making predictions on the test dataset, using the values of the explanatory variables in the data. We then calculate the RMSE using the errors between the predicted values and actual values of the testing data, and compare it to the RMSE of the original model built with the training data [@citeArnholt]. After using the model to make predictions on the testing data, it was found that the predicted values had a RMSE of 7.74. This is very similar to the original models RMSE based on the training data, which was 7.29. Since these values are very similar to one another, it means that our model did a good job at predicting values using the testing dataset, and thus it should also successfully be able to make predictions on other datasets, like for example future college basketball seasons. This means that the model has successfully been validated using the testing dataset, and we can now analyze what exactly was learned from our model.

@@

## Simple Rating System and its predictors

In this paper, we used several statistical methods in order to analyze a dataset containing team statistics on NCAA Division 1 men's basketball schools from the past five seasons. This was done to find out which specific statistics played an important role in how successful a team was. Through the use of several visualizations such as Figure \@ref(fig:scatters), it was evident that several advanced statistics had a clear relationship with a teams Simple Rating System (SRS), a value that determines how successful a team is based on the point differential of their games and the quality of their opposition. These advanced statistics were first checked for multicollinearity through the use of DAGs, seen in Figure \@ref(fig:dags), before being used to create several different linear regression models. Each model had a different combination of explanatory variables used to predict our response, SRS, and the models were tested against each other to determine which one was the best. This was done by comparing several goodness of fits tests across each of the models, and seeing which linear relationship best fit the used data.

As a result of these methods, it was determined that there were three team statistics that could be used to predict how successful a team would be. These three variables are a teams true shooting percentage, total rebounding percentage, and turnover percentage. Their inclusion in the model suggest that each of the team statistics play a large role in how successful a team is since together they all have a relationship with simple rating system. Teams that hit more of their shots and have a higher shooting percentage will score more points, increasing their simple rating system value. Simple rating system will also improve when a team has a higher rebounding percentage, since grabbing more rebounds than the opposition will simultaneously give a team more chances to score will giving the opposition less chances. Turnover percentage has a negative relationship with simple rating system, as teams with less turnovers than others will have possession of the ball more often and for longer, giving them more time and chances to score than their opposition.

While other statistics do also play a role in how good a team is, these statistics in particular play the biggest roles and have the strongest relationship with simple rating system, which is why they were the only variables included in the final model. In the presence of each other, these three statistics and their values will be the main reasons why a team is successful or not. Teams with high true shooting and total rebounding percentages, and low turnover percentages will be the most successful teams in the league, as their simple rating system values will be the highest. On the contrary teams with low true shooting and total rebounding percentages, and high turnover percentages will have low SRS values and struggle to succeed. So, our research question has been answered, and team success can be predicted using advanced team statistics.

## March Madness Brackets, Betting and Coaching

There are several pieces of information that can be learned and obtained from these findings. To recall, a big reason that this research on college basketball was completed is due to the popularity of the March Madness tournaments, and specifically the creation of brackets by many. The goal was to obtain information that would help people make better predictions when creating their bracket. Since we have learned that true shooting, total rebounding, and turnover percentages are the most important statistics in college basketball, fans can put a large emphasis on these statistics when making decisions in their brackets. Fans could simply substitute a teams values in the current season into the model formula described in \@ref(results) and calculate the teams expected simple rating system value. Then, they could choose their bracket based on which teams have the higher value of SRS. Considering the strength and success of the model, doing such a thing would result in a pretty good bracket as the stats should do a very good job at predicting success.

Of course, a fan doing such a thing is not very realistic, as there is little fun in this and many would not have access to the model, but just knowing what was found through the creation of the model would also be very helpful to fans in many other ways. For starters, year after year there are several crazy upsets in the March Madness tournament that shock the world as no one saw them coming [@citeGleeson]. However, using our findings may give fans a one up over other when it comes to predicting these upsets. If we look at the values of the statistics used as predictors in our model for teams who had poorer records in the season and are thus ranked lower in the tournament brackets, we could get insight on which teams are underanked and are more likely to upset better teams. For example, if a fan realizes a low ranked seed has a very good shooting percentage, rebounds the ball good, and does not turn the ball over, based on the model the chances of this team having success in the tournament and upsetting a higher ranked seed is quite high. This team may have gotten unlucky throughout the season which resulted in their poor record and rank, but based on the underlying important statistics we can realize that they should be a better team, and would be a good underdog pick in the bracket. The same goes for high ranked teams with poor values of these predictor statistics who got lucky in the season and won a lot of games, as realizing this would allow fans to see picking them to go far is a bad idea. Apart from predicting underdogs, if a fan is having a hard time picking between two teams in a specific match up and cannot decided which one to pick, they could turn to comparing the two teams true shooting, total rebound, and turnover percentages in order to come to a decision. If the two teams have equal shooting and turnover percentages but one team has a much higher rebounding percentage, the better rebounding team has a higher chnace of being successful based on the model, and thus this information would help the undecisive fan come to a conclusion and pick the better team.

For the same reasons the model and its findings are helpful to fans making brackets, it would also be very helpful to people who bet on games.  As of recently, sports betting has seen a massive surge of popularity due to the ease of betting online [@citeHudson]. Those who bet on college basketball games can get a huge monetary benefit by using the findings of this paper. By focusing in on teams values for the predictor statistics in the model, bettors can recognized which odds set by oddsmakers in Vegas are most beneficial to them. For example, if one team has very low odds of winning the game, but has good underlying statistics and the model predicts they should be a more sucessful team than the team with high odds of winning, then bettors could place money on the underdog team and have a very high chances of receiving a massive payday. It is very difficult for people to beat the oddsmakers in sports betting, but using the knowlegde obtained from this paper can give bettors an advantage that many people do not have.

Finally, although the findings were intended to be for fans, they could also be helpful to the coaches of college basketball teams who can or will play in March Madness.  Since we found that good shooting, good rebounding, and limited turnovers leads to team success, coaches can go and recruit players who will help their team obtain better results in these specific categories. They could built their team with players who shoot the ball well, are tall and have advantages rebounding, and are safe with the basketball to improve their simple ratig system value, as predicted by the model, and have more success. This could also be done by teaching their players how to improve in these areas both individually and as a team through game plans and practice drills. Using the findings in the paper could give coaches a huge benefit over others, increasing their chances at making their dreams come true and winning March Madness.

## Limitations

While the findings in this paper are mostly trustworthy and extremely helpful, there were several limitations that may have affected our results. As mentioned in \@ref(data), there are several biases, both racial and accidental, in the data that was used. As a result of scorekeepers and referees being prone to human error, and the overall ambiguity of many basketball statistics, it may have caused the data for some teams to be skewed. For example, if two scorekeepers in two different arenas have different ideas of what a turnover is, they will track the stat differently. As a result of this, one team who has a scorekeeper who tracks more turnovers relative to scorekeepers of other teams could have a turnover percentage much higher than it should be. This would mean our model predicts they should have a lower SRS and be a worse team even though though they do not turn the ball over and their scorekeeper just believes they do. This could occur for any other of the team statistics we looked at, meaning many values in the data are likely biased and skewed, which could have affected our models estimates and even which variables were included in the model. Also, referees are known to have racial biases, and tend to give more calls to players and teams of the same race as them. If a team is predominantly white and has a majority of white referees throughout the season, their team stats will be inflated, and the model will predict that the team is better than they are due to the biases. Due to this, people will use these findings and favour the team that benefited from the biases, which may work until the team gets a referee of a different race who does not help them out, and the teams actual abilities are exposed.

Apart from the biases, there were a few other limitations with the dataset that was used. As everyone is well aware of, the COVID-19 pandemic has had a large effect on nearly every aspect of life, and that includes college basketball. Over the past three seasons, hundreds of games had to be canceled because of the virus. 2020 saw the entire March Madness and many conference tournaments go unfinished as the virus first began to spread. It only got worse from here, as no Ivy League schools participated in the following season [@citeDiamond], and in only half of the most recent season over 70 schools had to shut down their basketball operations [@citeBorz]. Overall, as a result of all these cancellations, many games went unplayed and thus the sample size of our dataset, and the amount of total games played which are used to calculate values in the data, was much smaller than it would have been in normal seasons. The smaller and affected sample size may have led to our models coefficient estimations being slightly less accurate than they could have been. Finally, the dataset that was used only contained a certain amount of team statistics. There are many simple and advanced statistics out there that were not accounted for when creating our model, and thus a key predictor of simple rating system could have been overlooked. Despite this, the statistics analyzed were the main and most common advanced statistics, so this should not be too much of a worry.

## Next Steps

Thankfully, there are steps that can be taken in the future to enhance our findings and mitigate the effects of these limitations. To account for the biases in our dataset, we could only use a sample of the data that is more consistent throughout when it comes to scorekeeping and refereeing. For example, only games played on neutral sites could be taken into account to minimize the amount of scorekeeper bias. Also, to remove racial bias in the data, a sample of only games where the refereeing team was diversified and multiple races were equally represented. By doing such a thing, we will insure that all data is tracked with consistency and that no team is favoured over another.

Furthermore, as more seasons are played in the future, we will obtain more complete data that will give us a larger sample size. This means our dataset would no longer be affected as badly as it was due to the COVID-19 pandemic, assuming our world improves over time in regards to the virus. To further complement this, datasets from various other sources could be searched to find data on advanced statistics that were not included in our dataset. Together, this would give us a dataset that is much larger and contains many more variables, ensuring that our model estimates are as accurate as possible and that all variables that have a relationship with team success are actually included.
